skater: An R package for SNP-based Kinship Analysis, Testing, and Evaluation

### 2021-07-12 paper/paper.tex
compile paper and vignette

\access{Advance Access Publication Date:   }
\appnotes{Manuscript Category}

@@

\subtitle{Application Note}

\title[skater package for kinship analysis]{\texttt{skater}: An R
package for SNP-based Kinship Analysis and Benchmarking}

\author[FirstAuthorLastName \textit{et~al}.]{
Stephen D. Turner\,\textsuperscript{1},
V.P. Nagraj\,\textsuperscript{1},
Matthew Scholz\,\textsuperscript{1},
Shakeel Jessa\,\textsuperscript{1},
Carlos Acevedo\,\textsuperscript{1},
Jianye Ge\,\textsuperscript{2},
August Woener\,\textsuperscript{2},
Bruce Budowle\,\textsuperscript{2},
}

@@

\abstract{
\textbf{Motivation:} This section should specifically state the
scientific question within the context of the field of study.\\
\textbf{Results:} This section should summarize the scientific advance
or novel results of the study, and its impact on computational
biology.\\
\textbf{Availability:} This section should state software availability
if the paper focuses mainly on software development or on the
implementation of an algorithm. Examples are: `Freely available on the
web at XXX.' Website implemented in Perl, MySQL and Apache, with all
major browsers supported'; or `Source code and binaries freely available
for download at URL, implemented in C++ and supported on linux and MS
Windows'. The complete address (URL) should be given. If the manuscript
describes new software tools or the implementation of novel algorithms
the software must be freely available to non-commercial users. Authors
must also ensure that the software is available for a full TWO YEARS
following publication. The editors of Bioinformatics encourage authors
to make their source code available and, if possible, to provide access
through an open source license.\\
\textbf{Contact:}sturner@signaturescience.com\\
\textbf{Supplementary information:} Supplementary data are available at
Bioinformatics Online.}

@@

Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

@@

Describe the approach. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

Describe the approach. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

Describe the approach. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

Describe the approach. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

\section{Functions and usage}

Detailed methods. Lorem ipsum ad nauseum. Introduce your topic. Lorem
ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

Detailed methods. Lorem ipsum ad nauseum. Introduce your topic. Lorem
ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

Detailed methods. Lorem ipsum ad nauseum. Introduce your topic. Lorem
ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

\subsection{Pedigree processing}

Details for Method 1. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

Details for Method 1. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

Details for Method 1. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

Details for Method 1. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

@@

Details for Method 2. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

Details for Method 2. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

Details for Method 2. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

Details for Method 2. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

\subsection{Additional features and built-in data}

Details for Method 2. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

Details for Method 2. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

Details for Method 2. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

Details for Method 2. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad nauseum.
Introduce your topic. Lorem ipsum ad nauseum. Introduce your topic.
Lorem ipsum ad nauseum.

@@

Anything else? Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum
ad nauseum. Introduce your topic. Lorem ipsum ad nauseum. Introduce your
topic. Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad
nauseum.

Anything else? Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum
ad nauseum. Introduce your topic. Lorem ipsum ad nauseum. Introduce your
topic. Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad
nauseum.

Anything else? Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum
ad nauseum. Introduce your topic. Lorem ipsum ad nauseum. Introduce your
topic. Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad
nauseum.

Anything else? Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum
ad nauseum. Introduce your topic. Lorem ipsum ad nauseum. Introduce your
topic. Lorem ipsum ad nauseum. Introduce your topic. Lorem ipsum ad
nauseum.

\section*{Acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}

These should be included at the end of the text and not in footnotes.
Please ensure you acknowledge all sources of funding, see funding
section below.

Details of all funding sources for the work in question should be given
in a separate section entitled `Funding'. This should appear before the
`Acknowledgements' section.

@@

The following rules should be followed:

>>>>

\access{Advance Access Publication Date:   }
\appnotes{Applications Note}

@@

\subtitle{Genetics and Population Analysis}

\title[skater package for kinship analysis]{skater: An R package for
SNP-based Kinship Analysis, Testing, and Evaluation}

\author[FirstAuthorLastName \textit{et~al}.]{
Stephen D. Turner\,\textsuperscript{1}, V.P.
Nagraj\,\textsuperscript{1}, Matthew
Scholz\,\textsuperscript{1}, Shakeel Jessa\,\textsuperscript{1}, Carlos
Acevedo\,\textsuperscript{1}, Jianye Ge\,\textsuperscript{2}, August
Woener\,\textsuperscript{2}, Bruce Budowle\,\textsuperscript{2}
}

@@

\abstract{
\textbf{Motivation:} SNP-based kinship analysis with genome-wide
relationship estimation and IBD segment analysis methods produce results
that often require further downstream manipulation and analysis. A
software package with a consistent design and implementation that offers
this downstream analysis functionality is needed.\\
\textbf{Results:} Here we present the skater package for
\textbf{S}NP-based \textbf{k}inship \textbf{a}nalysis, \textbf{t}esting,
and \textbf{e}valuation as an \textbf{R} package. The skater package
contains a suite of well-documented tools for importing, parsing, and
analyzing pedigree data, performing relationship degree inference,
benchmarking relationship degree classification, and summarizing IBD
segment data.\\
\textbf{Availability:} The skater package is implemented as an R package
and is released under the MIT license at
https://github.com/signaturescience/skater. Documentation is available
at https://signaturescience.github.io/skater.\\
\textbf{Contact:}sturner@signaturescience.com\\
\textbf{Supplementary information:} Supplementary data are available at
Bioinformatics Online.}

@@

Inferring familial relationships between individuals using genetic data
is a common problem in population genetics, medical genetics, and
forensics. There are multiple approaches to estimating the relatedness
between samples, including genome-wide relatedness measures such as
those implemented in Plink \citep{purcell2007} or KING
\citep{manichaikul2010}, and methods that rely on identity by descent
(IBD) segment detection such as GERMLINE \citep{gusev2009}, hap-IBD
\citep{zhou2020}, and IBIS \citep{seidman2020}. Recent efforts have
focused on benchmarking these methods \citep{ramstetter2017, vries2021}
aided by tools for simulating pedigrees and genome-wide SNP data
\citep{caballero2019}. Analyzing results from genome-wide SNP-based
kinship analysis or comparing analyses to simulated data for
benchmarking have to this point required writing one-off analysis
functions or utility scripts that are rarely shared and/or poorly
documented. There is a need in the field for a well-documented software
package with a consistent design and API that contains functions to
assist with downstream manipulation, benchmarking, and analysis of
SNP-based kinship assessment methods. Here we present the skater R
package for \textbf{S}NP-based \textbf{k}inship \textbf{a}nalysis,
\textbf{t}esting, and \textbf{e}valuation.

@@

The skater package provides a tidyverse-friendly collection of analysis
and utility functions for \textbf{S}NP-based \textbf{k}inship
\textbf{a}nalysis, \textbf{t}esting, and \textbf{e}valuation as an
\textbf{R} package. Functions in the package include tools for
importing, parsing, and analyzing pedigree data, performing relationship
degree inference, benchmarking relationship degree classification, and
summarizing IBD segment data.

\subsection{Pedigree parsing, manipulation, and analysis}

The skater package has several functions for importing, parsing, and
analyzing pedigree data. Pedigrees define familial relationships in a
hierarchical structure. Many genomics tools for working with pedigrees
start with a .fam file, which is a tabular format with one row per
individual and columns for unique IDs of the mother, father, and the
family unit. The skater package contains the functions
\texttt{read\_fam()} to read in a PLINK-formatted .fam file, and another
function \texttt{fam2ped()} to convert these into an internal pedigree
object as a nested tibble with one row per family. Further functions
such as \texttt{plot\_pedigree()} produce a multi-page PDF drawing a
diagram of the pedigree for each family, while \texttt{ped2kinpair()}
produces a pairwise list of relationships between all individuals in the
data with the expected kinship coefficients for each pair.

\subsection{Relationship degree inference and benchmarking}

The skater package includes functions to translate kinship coefficients
to relationship degrees. The kinship coefficients could come from
\texttt{ped2kinpair()} or other kinship estimation software.

The \texttt{dibble()} function creates a \textbf{d}egree
\textbf{i}nference \textbf{t}ibble, with degrees up to the specified
maximum degree resolution, expected kinship coefficient, and lower and
upper inference ranges as defined in \citet{manichaikul2010}. The
\texttt{kin2degree()} function infers the relationship degree given a
kinship coefficient and a maximum degree resolution (e.g., 7th-degree
relatives) up to which anything more distant is treated as unrelated.

Once estimated kinship is converted to degree, it may be of interest to
compare the inferred degree to truth. When aggregated over many
relationships and inferences, this can help benchmark performance of a
particular kinship analysis method. The skater package adapts a
\texttt{confusion\_matrix()} function from \citet{clark2021} to provide
standard contingency table metrics (e.g.~sensitivity, specificity, PPV,
precision, recall, F1, etc.) with a new reciprocal RMSE (R-RMSE) metric.
The R-RMSE metric is defined more thoroughly in the skater package
vignette (see Supplementary Material), and is a superior measure to
classification accuracy when benchmarking relationship degree
estimation. Taking the reciprocal of the target and predicted degree in
a typical RMSE calculation results in larger penalties for more
egregious misclassifications (e.g., classifying a first-degree relative
pair as second degree) than misclassifications at more distant
relationships (e.g., misclassifying a fourth-degree relative pair as
fifth-degree).

@@

Tools such as hap-IBD \citep{zhou2020}, and IBIS \citep{seidman2020}
detect shared IBD segments between individuals. The skater package
includes functionality to take those IBD segments, compute shared
genomic centimorgan (cM) length, and convert that shared cM to a kinship
coefficient. In addition to inferred segments, these functions can
estimate ``truth'' kinship from simulated IBD segments
\citep{caballero2019}. The \texttt{read\_ibd()} function reads pairwise
IBD segments from IBD inference tools and from simulated IBD segments.
The \texttt{read\_map()} function reads in genetic map in a standard
format which is required to translate the total centimorgans shared IBD
to a kinship coefficient using the \texttt{ibd2kin()} function.

@@

The skater R package provides a thoroughly-documented software package
for common data import, manipulation, and analysis tasks typically
encountered when working with common SNP-based kinship analysis tools.
All package functions are internally documented with examples, and the
package contains a vignette demonstrating usage, inputs, outputs, and
interpretation of all key functions (see Supplementary Material). The
package contains internal tests which are automatically run with
continuous integration via GitHub Actions whenever the package code is
updated. The skater package is permissively licensed (MIT) and is easily
extensible to accommodate outputs from new genome-wide relatedness and
IBD segment methods as they become available.

@@

\textbf{UNT FOLKS: NEED YOUR INPUT HERE!} The following rules should be
followed:
